Missing data: Current practice in football research and recommendations for improvement

### 2020-10-15 paper/paper.Rmd
start to bring in mismatches, fix spelling, fix the four panel figure

```{r read-data}
gs_football <- readr::read_rds(here::here("data",
                           "gs-football.rds"))
```

@@

```{r count-words}
wordcountaddin::text_stats()
```

@@

For clarity, we describe an example of missingness. An Australian Football data collector 
was taking lunch during the third quarter, so no data was recorded for this time period. This
would be considered missing data. Compare this to the game being cancelled due to a pandemic. 
There is no missing data because there is no data to be observed. Similiarly, if a study screens out individuals who do not meet inclusion criteria, this is not a missing value, because there is no intent for these values to be observed or measured.

@@

Broadly, there are two types of missing values in data: implicit and
explicit missings. Explicit values are missing, but recorded; whereas
with implicit missinges, their presence is implied based on other
information in the data. For example, in  Table \@ref(tab:implicit-missings), player Koenen has
missing values for quarters two and four. Sometimes values like these
can be logically imputed, as it might be known that these values are
recorded as 0 in this format, rather than NA. Other places missing data
can arise include: through joins when merging data without corresponding
values, surveys, an inability to collect a biological sample (e.g.,
venous blood), equipment malfunction, failure, or not being worn.

@@

# Simulate dataset
d <- get_afltables_stats(start_date = "2020-08-01", end_date = "2020-09-01")

data_afl <- d %>% select(ID, Round, Playing.for, Venue)

# Simulate injury and sprint variables
injury <- rbinom(n = nrow(d), 1, prob = 0.1)
sprints <- rpois(n = nrow(d), lambda = 100)
sprint_distance <- truncnorm::rtruncnorm(n = nrow(d), a = 0, b = Inf, mean = 2000, sd = 500) # Check if this is reasonable (mean 2 km per game)

@@

# Put the dataset together
data_plot <- cbind(data_afl, data_with_missing)

@@

@@

aq_vis_miss <- vis_miss(data_plot) #+ 
  #theme(text = element_text(size = 14))

aq_fct <- gg_miss_fct(data_plot, team) + 
  #theme(text = element_text(size = 14)) +
  labs(x = "Team", y = "Variable")

gg_miss_scatter <- ggplot(data_plot,
       aes(x = sprint_distance,
           y = sprints)) +
  geom_miss_point(size = 2) + 
  theme_minimal() +
  #theme(text = element_text(size = 14)) +
  labs(x = "Total sprint distance (m)", y = "Number of sprints (count)", colour = "Missing")

aq_miss_var <- gg_miss_var(data_plot)

# Grid plots
cowplot::plot_grid(aq_vis_miss, aq_fct, gg_miss_scatter, aq_miss_var, 
                   nrow = 2,
                   ncol = 2,
                   align = "hv",
                   axis = "tblr",
                   label_size = 20,
                   labels = "AUTO")
```

@@

Methods that provide multiple imputed values for the outcome, such as multiple imputation, or using a Bayesian framework that generates a distribution of posterior values, can be appropriate for imputed the outcome as they provide some sence of what the uncertainty is. For discussion on imputing the response value, see ... [@refs].

@@

Our survey of articles on football related topics
showed that current practice of reporting missing data is poor, 
with only about one in 20 studies reporting missingness.
We speculate that this result is representative of practice in the 
wider sports science literature. The extremely low reporting of missing 
data could suggest that this aspect of anlaysis receives little attention. 
To address this issue and assist researchers,
we have provided recommendations for reporting and exploring missing data.
Researchers should consider these recommendations, and pay greater
attention to missing data and its influence on research results.

>>>>

```{r read-data, message = FALSE}
gs_football <- readr::read_rds(here::here("data",
                           "gs-football.rds"))
```

```{r mismatch}
# Now we bring the third and final review 
gs_mismatches <- readr::read_csv(here::here("data",
"missing-data-mismatches.csv"))

gs_mismatch_pivot <- gs_mismatches %>% 
  pivot_longer(cols = c(reviewer_1,reviewer_2, third_check),
               names_to = "reviewer",
               values_to = "decision") %>% 
  pivot_wider(names_from = variable,
              values_from = decision)
```

@@

```{r count-words, message = FALSE, warning = FALSE}
wordcountaddin::text_stats()
```

@@

For clarity, we describe an example of missingness. An Australian Football data collector 
was taking lunch during the third quarter, so no data was recorded for this time period. This
would be considered missing data. Compare this to the game being cancelled due to a pandemic. 
There is no missing data because there is no data to be observed. Similarly, if a study screens out individuals who do not meet inclusion criteria, this is not a missing value, because there is no intent for these values to be observed or measured.

@@

Broadly, there are two types of missing values in data: implicit and
explicit missings. Explicit values are missing, but recorded; whereas
information in the data. For example, in  Table \@ref(tab:implicit-missings), player Koenen has
missing values for quarters two and four. Sometimes values like these
can be logically imputed, as it might be known that these values are
recorded as 0 in this format, rather than NA. Other places missing data
can arise include: through joins when merging data without corresponding
values, surveys, an inability to collect a biological sample (e.g.,
venous blood), equipment malfunction, failure, or not being worn.

@@

# Simulate dataset
# NOTE: fix this later in the drake plan
# d <- get_afltables_stats(start_date = "2020-08-01", end_date = "2020-09-01")
# write_rds(d, here::here("data", "afl-data.rds"))
d <- readr::read_rds(here::here("data", "afl-data.rds"))
afl_data <- d %>% select(ID, Round, Playing.for, Venue)

# Simulate injury and sprint variables
injury <- rbinom(n = nrow(d), 1, prob = 0.1)
sprints <- rpois(n = nrow(d), lambda = 100)
sprint_distance <- truncnorm::rtruncnorm(n = nrow(afl_data), 
                                         a = 0, 
                                         b = Inf, 
                                         mean = 2000, 
                                         sd = 500) # Check if this is reasonable (mean 2 km per game)

@@

# Put the dataset together
data_plot <- cbind(afl_data, data_with_missing)

@@

```{r}
data_plot_clean <- data_plot %>% 
  mutate(team = recode(team, 
                        "Gold Coast" = "GC",
                        "Greater Western Sydney" = "GWS",
                        "Fremantle" = "FRE",
                        "Collingwood" = "COL",
                        "Port Adelaide" = "PA",
                        "Western Bulldogs" = "WB",
                        "Richmond" = "RIC",
                        "Brisbane Lions" = "BL",
                        "Geelong" = "GEE",
                        "North Melbourne" = "NM",
                        "Adelaide" = "ADE",
                        "Melbourne" = "MEL",
                        "Sydney" = "SYD",
                        "St Kilda" = "STK",
                        "Essendon" = "ESS",
                        "West Coast" = "WC",
                        "Carlton" = "CAR",
                        "Hawthorn" = "HAW")) %>% 
  as_tibble() %>% 
  rename(sprint_dist = sprint_distance)

@@

aq_vis_miss <- vis_miss(data_plot_clean) + labs(tag = "A")
  #theme(text = element_text(size = 14))

aq_fct <- gg_miss_fct(data_plot_clean, team) + 
  #theme(text = element_text(size = 14)) +
  labs(x = "Team", y = "Variable", tag = "B") +
  theme(axis.text.x = element_text(size = 5))

gg_miss_scatter <- ggplot(data_plot_clean,
       aes(x = sprint_distance,
           y = sprints)) +
  geom_miss_point(size = 2) + 
  theme_minimal() +
  #theme(text = element_text(size = 14)) +
  labs(x = "Total sprint distance (m)", 
       y = "Number of sprints (count)", 
       colour = "",
       tag = "C")

aq_miss_var <- gg_miss_var(data_plot_clean) + labs(tag = "D")
```

```{r patch-2, fig.height = 10}
(aq_vis_miss + aq_fct) / (gg_miss_scatter + aq_miss_var) & 
  theme(aspect.ratio = 1,
        legend.position = "bottom")
```

@@

Methods that provide multiple imputed values for the outcome, such as multiple imputation, or using a Bayesian framework that generates a distribution of posterior values, can be appropriate for imputed the outcome as they provide some sense of what the uncertainty is. For discussion on imputing the response value, see ... [@refs].

@@

Our survey of articles on football related topics
showed that current practice of reporting missing data is poor, 
with only about one in 20 studies reporting missingness.
We speculate that this result is representative of practice in the 
wider sports science literature. The extremely low reporting of missing 
data could suggest that this aspect of analysis receives little attention. 
To address this issue and assist researchers,
we have provided recommendations for reporting and exploring missing data.
Researchers should consider these recommendations, and pay greater
attention to missing data and its influence on research results.
